Over Twenty Years of Virtual Machine Development and Debugging Through Simulation

### 2018-08-15 paper.tex
Added more figures

@@

\begin{figure*}[bth!]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

Bindings to processor simulator libraires (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated. Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction. Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation. Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.

@@

IN-IMAGE COMP FIG?

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}.

The JIT basically generates a sequence of machine instructions for a given pattern of bytecodes. We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method.

>>>>

\begin{figure*}[ht!]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/SimulatedMemory}
		\caption{Runtime Memory and Simulated Counter-Part}
		\label{fig:SimulatedMemory}
\end{figure*}

@@

@@

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated. Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction. Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation. Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.

To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code, calls the bound processor simulator to disassemble the code and decorate the disassembly with Smalltalk specific information. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

\begin{figure}[ht!]
		\centering
		\includegraphics[width=\linewidth]{figures/InImageCompiler}
		\caption{In-image compilation}
		\label{fig:InImageCompiler}
\end{figure}

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.
